---
title: "Development actions history"
output: html_document
editor_options: 
  chunk_output_type: console
---

All commands that you use to use when developing packages...

# First time just after creating the project

- Fill the following chunk to create the DESCRIPTION of your package

```{r description, eval=FALSE}
# Describe your package
fusen::fill_description(
  pkg = here::here(),
  fields = list(
    Title = "Spatial Point Pattern Analysis with FDA",
    Version = "0.1",
    Description = "Function to analyse TCGA dataset",
    `Authors@R` = c(
      person("Valerio", "Reffo", email = "valerio.reffo@gmail.com", role = c("aut", "cre"), comment = c(ORCID = "0000-0000-0000-0000"))
    )
  )
, overwrite = TRUE)
# Define License with use_*_license()
usethis::use_gpl3_license()
```

# Start using git

Not required, since we are already on a repository.

```{r, eval=FALSE}
#usethis::use_git()
# Deal with classical files to ignore
#usethis::git_vaccinate()
# Use main for primary branch
#usethis::git_default_branch_rename()
```

# Set extra sources of documentation

```{r, eval=FALSE}
# Install a first time
remotes::install_local()
# README
usethis::use_readme_rmd()
# Code of Conduct
usethis::use_code_of_conduct("contact@fake.com")
# NEWS
usethis::use_news_md()
```

**From now, you will need to "inflate" your package at least once to be able to use the following commands. Let's go to your flat template, and come back here later if/when needed.**

# Inflate flat_Lcross_pipeline.Rmd

Start by inflating `dev/flat_Lcross_pipeline.qmd`:

```{r development-inflate, eval=FALSE}
# Run but keep eval=FALSE to avoid infinite loop
# Execute in the console directly
fusen::inflate(flat_file = "dev/flat_Lcross_pipeline.qmd",
               vignette_name = "dev-Lcross_pipeline",
               overwrite = T)
```


This will put everything in the right place to get started. We then continue by
adding the fda pipeline.

```{r development-inflate, eval=FALSE}
# Run but keep eval=FALSE to avoid infinite loop
# Execute in the console directly
fusen::inflate(flat_file = "dev/flat_fda_pipeline.qmd",
               vignette_name = "dev-fda_pipeline",
               overwrite = T)
```

# Including datasets

The datasets created in the `save_data.Rmd` have still to be saved into the
`data/` folder of the package. This will make the datasets available using
`data("commune_level_data")`. But the package has to be inflated once at least
before being able to save the data and document it. We are going to knit the
vignette generated from `save_data.Rmd` for this. This vignette contains all the
calls needed to save the data. Then, we are going to use `usethis::use_data()`
to save the datasets. We knit the vignette into a temporary file using `tempfile()`
because we’re not interested in the output itself, but simply want the datasets
to be generated. When using `knitr::knit()`, every object that is created in the 
knitted vignette would be made available in the global environment. In this case,
knitting the vinette would make `commune_level_data`, `country_level_data` and
`communes` available in the global environment, which in turn would make it possible
to save using `usethis::use_data(commune_level_data)`. But we don’t want to clutter
the global environment with these datasets, we only want to save them.

So to avoid cluttering the global environment, we run the calls inside `local()`.
`local()` runs code in a temporary environment which is then discarded. This makes
it possible to turn this operation pure:

```{r}
local({
  knitr::knit("vignettes/dev-save_data.Rmd", output = tempfile())
  usethis::use_data(commune_level_data, overwrite = TRUE)
  usethis::use_data(country_level_data, overwrite = TRUE)
  usethis::use_data(communes, overwrite = TRUE)
})

```

The first call, to `knitr::knit()` knits the vignette, and makes the datasets
available to the local environment. Inside this local environment it is now possible
to call `usethis::use_data()` to save each data set in turn. The datasets are
now save inside the `data/` folder of the package, and are thus available to users
of the package using `data("dataset_name")`.

We still need to document the datasets. For this, we are going to inflate
another `.Rmd` file. Take a look at `dev/data_doc.Rmd`. This vignette documents the
data sets when we inflate it. Since this vignette is only useful for documenting the 
datasets, we don’t want to include it as a vignette in the package, so we set its name
to `NA`.

```{r}
fusen::inflate(flat_file = "dev/data_doc.Rmd",
               vignette_name = NA,
               overwrite = T)
```


# Package development tools
## Use once

Only use the commands that you need. You can safely delete those that you never run.

```{r, eval=FALSE}
# Pipe
usethis::use_pipe()

# package-level documentation
usethis::use_package_doc()

# GitHub
# Add your credentials for GitHub
gitcreds::gitcreds_set()
# Send your project to a new GitHub project
usethis::use_github()

# Set Continuous Integration
# _GitHub
usethis::use_github_action_check_standard()
usethis::use_github_action("pkgdown")
usethis::use_github_action("test-coverage")
# _GitLab
gitlabr::use_gitlab_ci(type = "check-coverage-pkgdown")

# Add new flat template
fusen::add_flat_template("add")

# Remove renv from the package
usethis::use_build_ignore("renv")
usethis::use_build_ignore("renv.lock")
```

## Use everytime needed

```{r}
# Simulate package installation
pkgload::load_all()

# Generate documentation and deal with dependencies
attachment::att_amend_desc()

# Check the package
devtools::check()
```


# Share the package

```{r}
# set and try pkgdown documentation website
usethis::use_pkgdown()
pkgdown::build_site()

# build the tar.gz with vignettes to share with others
devtools::build(vignettes = TRUE)
```

