---
title: "Functional Data Analysis and Clustering on L-cross Functions"
author: "Valerio Reffo"
date: "`r Sys.Date()`"
format: html
params:
  input_file: "/projects/shared/TCGA_data/results/combined_Lcross_KDE.csv"
---

## Setup and Libraries


```{r}
#| label: setup
#| message: false
#| warning: false

library(data.table)
library(parallel)
library(fda)
library(ggplot2)
library(dplyr)
library(tidyr)
```

## Data Loading

We read the combined $L$-cross Kernel Density Estimation (KDE) data from the specified path in the document parameters.

```{r}
#| label: load-data

message("Reading data from: ", params$input_file)
dt <- fread(params$input_file)
message("Successfully loaded data with ", nrow(dt), " rows and ", ncol(dt), " columns.")
```

## Functional Data Smoothing

We apply the `fsmooth` function to convert the raw data points into a continuous functional data object.

```{r}
#| label: fun-fsmooth

# safer smoothing wrapper that avoids using every observed r as a knot and
# provides sensible defaults for nbasis and lambda plus fallbacks on error
fsmooth <- function(
  dt,
  M = 6,                  # spline order (norder)
  genLfun.fd = 4,         # Lfdobj for fdPar
  centrata = FALSE,       # center by r before smoothing
  nbasis = NULL,          # allow user to set number of basis functions
  breaks = NULL,          # alternative: provide a coarser breaks vector
  lambda = NULL,          # smoothing penalty; if NULL a default is chosen
  rangeval = NULL,        # override range(r) if desired
  verbose = FALSE,
  ...
) {
  # 1. Robustly extract r
  if ("r_value" %in% names(dt)) {
    r <- dt$r_value
  } else {
    if (verbose) message("Warning: 'r_value' column not found; using first column as r")
    r <- dt[[1]]
  }

  # 2. Extract functional matrix (all columns after the first)
  mat_fun <- as.matrix(dt[, -1, with = FALSE])
  if (is.null(colnames(mat_fun))) {
    colnames(mat_fun) <- paste0("S", seq_len(ncol(mat_fun)))
  }

  # 3. Optional centering
  if (isTRUE(centrata)) {
    mat_fun <- sweep(mat_fun, MARGIN = 1, STATS = r, FUN = "-")
  }

  # 4. Determine range
  if (is.null(rangeval)) rangeval <- range(r, na.rm = TRUE)

  # 5. Choose nbasis sensibly if not provided
  n_r <- length(r)
  if (!is.null(breaks)) {
    # If the user supplied breaks, follow fda convention
    nbasis_calc <- length(breaks) + M - 2
    if (verbose) message("Using user-supplied breaks; nbasis computed as: ", nbasis_calc)
    genbasis <- create.bspline.basis(rangeval = rangeval, nbasis = nbasis_calc, norder = M, breaks = breaks)
  } else {
    if (!is.null(nbasis)) {
      nbasis_calc <- nbasis
    } else {
      # default heuristic: up to 30 bases but not larger than roughly n_r/4
      nbasis_calc <- min(30, max(M + 1, floor(n_r / 4)))
    }
    if (nbasis_calc >= n_r) {
      if (verbose) message("nbasis (", nbasis_calc, ") >= number of observations (", n_r, "); reducing nbasis to avoid overparameterization")
      nbasis_calc <- max(M + 1, n_r - 1)
    }
    genbasis <- create.bspline.basis(rangeval = rangeval, nbasis = nbasis_calc, norder = M)
  }

  # 6. Choose lambda if not provided (avoid near-interpolation)
  if (is.null(lambda)) {
    # A conservative default that provides some smoothing; users may increase
    # but avoid values like 1e-11 which effectively disable penalization
    lambda_val <- 1e-6
  } else {
    lambda_val <- lambda
  }

  genfdPar <- fdPar(genbasis, genLfun.fd, lambda = lambda_val)

  # 7. Try smoothing with fallback strategies on error
  genfdSmooth <- tryCatch(
    smooth.basis(argvals = r, y = mat_fun, fdParobj = genfdPar, ...),
    error = function(e) {
      if (verbose) message("smooth.basis failed: ", conditionMessage(e), "; retrying with larger lambda and Lfdobj=2")
      # fallback: increase penalty and use Lfdobj=2
      tryCatch(
        {
          genfdPar2 <- fdPar(genbasis, Lfdobj = 2, lambda = max(lambda_val, 1e-4))
          smooth.basis(argvals = r, y = mat_fun, fdParobj = genfdPar2, ...)
        },
        error = function(e2) {
          stop("Both primary and fallback smoothing attempts failed: ", conditionMessage(e2))
        }
      )
    }
  )

  Genfun.fd <- genfdSmooth$fd

  # 8. Set names defensively
  fdn_dist <- paste("Distanza (da 0 a", max(r, na.rm = TRUE), "micron)")
  fdnames <- list(fdn_dist,
                  "Sample" = colnames(mat_fun),
                  "Funzione")
  Genfun.fd$fdnames <- fdnames

  # 9. Basic sanity checks
  if (!is.null(Genfun.fd$coefs)) {
    n_coefs <- ncol(Genfun.fd$coefs)
    if (ncol(mat_fun) != n_coefs) {
      if (verbose) message("Warning: number of functions in mat_fun (", ncol(mat_fun), ") != number of fd coefficient columns (", n_coefs, ")")
    } else {
      if (verbose) message("Finished smoothing: nbasis=", genbasis$nbasis, ", lambda=", lambda_val)
    }
  } else {
    if (verbose) message("Finished smoothing but fd$coefs is NULL")
  }

  return(list(fd = Genfun.fd, fdPar = genfdPar, basis = genbasis, mat = mat_fun, r = r))
}
```


```{r}
#| label: fda-smoothing

message("Applying fsmooth function...")
smoothed_Lfun <- fsmooth(dt, M = 6, genLfun.fd = 4, centrata = FALSE)
```

## Functional Principal Component Analysis (fPCA)

We perform Functional PCA on the smoothed $L$-functions. Given the number of basis functions ($\text{length}(r) = 70$) and $M = 6$, the total number of basis functions is calculated as $K = 70 + 6 - 2 = 74$.

```{r}
#| label: fpca-analysis

# Run fPCA
fpca <- pca.fd(smoothed_Lfun$fd, nharm = 70, smoothed_Lfun$fdPar)

# Assign sample names (excluding the 'r_values' column) to the PCA scores
rownames(fpca$scores) <- colnames(dt)[-1]
```

## Unsupervised Clustering

We apply two different clustering techniques on specific dimensions of the fPCA scores (components 1 and 60) to classify the spatial patterns into 6 distinct groups.

### K-means & Hierarchical Clustering

```{r}
#| label: clustering

# Set seed for reproducibility
set.seed(123)

# K-means clustering
fpca_km <- kmeans(fpca$scores[, c(1, 60)], centers = 6)

# Hierarchical clustering (Ward's method)
fpca_hc <- hclust(dist(fpca$scores[, c(1, 60)]), method = "ward.D2")
memb_hc <- cutree(fpca_hc, k = 6)
```

### Cluster Distribution Comparison

Below are the distribution tables comparing the cluster assignments between both methods.

```{r}
#| label: cluster-tables

message("K-means cluster sizes:")
table(fpca_km$cluster)

message("Hierarchical clustering cluster sizes:")
table(memb_hc)
```

## Visualizations

### Smoothed L-functions Plot

We visualize the smoothed $L$-functions, color-coded by their K-means cluster assignments.

Function to plot the smoothed Lcross functions

```{r}
#| label: fun-plot_smoothed_L
plot_smoothed_Lfun <- function(smoothed_obj, clusters_vec, mark_i = "Mark 1", mark_j = "Mark 2", center_plot = TRUE) {
  
  # 1. Extract the distance vector (r)
  r_vec <- smoothed_obj$r
  
  # 2. Evaluate the functional data object (fd) to get smoothed values
  # This creates a matrix where rows are 'r' distances and columns are samples
  smooth_mat <- eval.fd(r_vec, smoothed_obj$fd)
  
  # 3. Convert the matrix to a data.frame and prepare for ggplot
  df_smooth <- as.data.frame(smooth_mat)
  df_smooth$r <- r_vec
  
  # Pivot to long format
  df_long <- pivot_longer(
    df_smooth,
    cols = -r,
    names_to = "Slide",
    values_to = "L_value"
  )
  
  # 4. Center the L-function (L(r) - r) if requested 
  # (Since you ran fsmooth with centrata = FALSE, we can center it visually here)
  if (center_plot) {
    df_long <- df_long |> mutate(L_plot_value = L_value - r)
  } else {
    df_long <- df_long |> mutate(L_plot_value = L_value)
  }
  
  # 5. Prepare the cluster metadata from the named vector (memb_hc)
  cluster_df <- data.frame(
    Slide = names(clusters_vec),
    cluster = as.factor(clusters_vec)
  )
  
  # 6. Join the functional data with the cluster assignments
  df_final <- df_long |> 
    left_join(cluster_df, by = "Slide")
  
  # 7. Build the ggplot
  p <- ggplot(df_final, aes(x = r, y = L_plot_value, group = Slide, color = cluster)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    geom_line(alpha = 0.7, linewidth = 0.6) + 
    theme_minimal() +
    labs(
      title = paste0("Smoothed L-cross Function: ", mark_i, " vs ", mark_j),
      subtitle = ifelse(center_plot, 
                        "Centered L-function (L(r) - r). Positive = Attraction, Negative = Repulsion", 
                        "L-function L(r)"),
      x = "Radius (r)",
      y = ifelse(center_plot, "L(r) - r", "L(r)"),
      color = "Cluster"
    ) +
    theme(legend.position = "right")
  
  return(p)
}
```


```{r}
#| label: plot-l-functions
#| fig-width: 8
#| fig-height: 6

plot_smoothed_Lfun(
  smoothed_obj = smoothed_Lfun, 
  clusters_vec = fpca_km$cluster, 
  mark_i = "neoplastic",  
  mark_j = "stromal",
  center_plot = TRUE      # Subtracts 'r' since centrata = FALSE in fsmooth
)

```

### PC Scores Comparison

Finally, we project and compare the sample distributions in the reduced PCA space using both clustering results.

```{r}
#| label: plot-pc-scores
#| fig-width: 10
#| fig-height: 5

# Set up side-by-side plotting layout
par(mfrow = c(1, 2))

# K-means plot
plot(fpca$scores[, 1], fpca$scores[, 2], col = fpca_km$cluster, pch = 3, main = "K-means")
abline(v = 0, lty = "dashed", col = "grey")
abline(h = 0, lty = "dashed", col = "grey")

# Hierarchical clustering plot
plot(fpca$scores[, 1], fpca$scores[, 2], col = memb_hc, pch = 3, main = "Ward.D2")
abline(v = 0, lty = "dashed", col = "grey")
abline(h = 0, lty = "dashed", col = "grey")

# Reset layout
par(mfrow = c(1, 1))
```