---
title: "Conversion of TCGA from adatas to Lcross functions"
author: "Valerio Reffo"
format: html
date: "`r Sys.Date()`"
---

```{r}
#| label: development
#| include: FALSE

library(testthat)
```


## Setup and Libraries

```{r}
#| label: setup
#| message: false
#| warning: false

library(spatstat)
library(imager)
library(anndataR)
library(parallel)
library(spatstat)
library(rhdf5)
```

I'm going to read all the sample adatas and the masks and add them togheter to create point process objects, then I'll reduce the dimensionality of the pp objects so I can estimate the Lcross function between two cell types of interest, by now they are Neoplastic and Stromal cells.

## Path Configuration

Dynamically configure paths for input, output, and masks based on document parameters.

```{r}
#| label: paths-handling

# Directories
sample_dir <-  "/projects/shared/TCGA_data/h5ad/TCGA-TS-A7P8-01Z-00-DX1.25E9AFA9-6C12-4F18-BE79-2F4F19887ECD.h5ad.gz"
save_dir <-  "/projects/shared/TCGA_data/results/test"

# Generate the save path
file_name <- sub("\\.h5ad\\.gz$", ".rds", basename(sample_dir))
save_path <- file.path(save_dir, file_name)

# Configure path for the mask (.png)
mask_path <- sub("/h5ad/", "/mask/", sample_dir)
mask_path <- sub("\\.h5ad\\.gz$", ".png", mask_path)
```

## Data Loading and PPP Conversion

Load the mask image and convert the spatial annotation file into a Point Pattern (`ppp`) object compatible with `spatstat`.

The function need the h5ad file path, the mask as an image and combine them together based on magnification level, 
it's needed to specify the column of the h5ad with the cell type information, and the name of the spatial coordinate, by default named "x" and "y". 

```{r}
#| label: function-ann2ppp
#' ann2ppp Read a h5ad file and convert it to a point process object from spatstat library
#' 
#' @param data_path path to the h5ad file
#' @param mask Optional: img object to be converted in the window of the point process class object
#' @param magnif level of magnification of the original H&E image, can be 20x or 40x
#' @param spat_coords_name names of coordinate of the centroids in the h5ad file
#' @param marks_col name of the .obs vector with the names of the marks (usually the cell types names)
#' @importFrom anndataR read_h5ad
#' @importFrom spatstat.geom owin affine ppp
#' @return An object of class "ppp"
#' @export
ann2ppp <- function(data_path, mask = NULL, magnif, spat_coords_name = c("x", "y"), marks_col = "type") {
  
  # magnif is the value of the magnification parameter in the metadata (can be 20 or 40 (20x or 40x respectively)) 
  # it's probably an error if it's noted as 20x.
  
  # 1. Read AnnData
  ann <- read_h5ad(data_path)
  
  # 2. Extract spatial coordinates
  spat <- ann$obsm$spatial
  x_coords <- spat[, 1]
  y_coords <- spat[, 2]
  
  # 3. Extract marks (e.g., cell types)
  marks_data <- NULL
  if (marks_col %in% colnames(ann$obs)) {
    marks_data <- as.factor(ann$obs[[marks_col]])
  }
  
  # 4. Define the window (owin)
  if (!is.null(mask)) {
    # Extract the 2D matrix from the imager cimg object (dropping z and c dimensions).
    # imager keeps 'x' as rows and 'y' as columns, but spatstat's owin mask 
    # expects 'y' as rows and 'x' as columns, so we transpose it.
    # We also convert it to a logical matrix (TRUE for foreground > 0)
    mask_mat <- t(mask[, , 1, 1] > 0.5) 

    # Create an owin using the image mask and the ranges of the spatial coordinates
    win_thumb <- owin(
      xrange = c(0, dim(mask)[1]), 
      yrange = c(0, dim(mask)[2]), 
      mask = mask_mat
    )
    # C. Calculate exact scale
    exact_scale <- magnif / 1.25
    
    # D. Transform the window in the full resolution scale of the coordinates
    final_win <- affine(win_thumb, mat = diag(c(exact_scale, exact_scale)))
    
  } else {
    # Create a basic bounding box window if no mask is provided
    final_win <- owin(
      xrange = range(x_coords, na.rm = TRUE), 
      yrange = range(y_coords, na.rm = TRUE)
    )
  }
  
  # 5. Build the PPP object
  p <- ppp(
    x = x_coords, 
    y = y_coords, 
    window = final_win,
    marks = marks_data
  )
  
  return(p)
}

```

Read the mask as an image

```{r}
#| label: mask-import

img <- imager::load.image(mask_path)
```

Create the point process object

```{r}
AppMag <-  40

pp_object <- ann2ppp(sample_dir, mask = img, magnif = AppMag)
```

## Downsampling (KDE)

To prevent memory overflow, we apply a Kernel Density Estimation (KDE) based downsampling before computing the `Lcross` function between Neoplastic and Stromal cells.

Function to downsample based on the Kernel Density Estimation, it needs a pp objects, the marks name which is a vector of dim 2 and a target number of point to which downsample. It's possible to personalized the density.ppp function

```{r}
#| label: function-downsample_KDE
#' downsample_KDE Function to downsample based on the Kernel Density Estimation.
#' 
#' It needs a pp objects, the marks name which is a vector of dim 2 and a target number of point to which downsample. 
#' It's possible to personalized the density.ppp function
#' 
#' @param ppp_obj object of class "ppp" from spatstat
#' @param marks_name vector of dim 2 with the names of the cell types to apply the function to
#' @param n_target number of point in the ppp object to downsample
#' @param ... additional specification for the density.ppp function
#' @importFrom parallel mclapply detectCores
#' @importFrom spatstat.geom marks split.ppp npoints lookup.im coords superimpose marks<-
#' @importFrom spatstat.explore density.ppp bw.ppl
#' @return An object of class "ppp" with n_target points for each marks_name
#' @export
downsample_KDE <- function(ppp_obj, marks_name, n_target, ...) {
  
  # 1. Subset to the specified marks
  pp_neo <- subset(ppp_obj, marks %in% marks_name)
  marks(pp_neo) <- factor(marks(pp_neo))
  
  # 2. Split into a named list based on the marks
  pp_split <- split.ppp(pp_neo)
  
  # 3. Process each split in parallel
  sampled_list <- mclapply(pp_split, function(pp) {
    
    n_available <- npoints(pp)
    
    # Safely handle empty patterns
    if (n_available == 0) return(pp[integer(0)])
    
    # Skip KDE if we don't have enough points to downsample
    if (n_available <= n_target) return(pp)
    
    # Calculate KDE safely with error handling for bandwidth selection
    kern <- tryCatch({
      density.ppp(pp, at="pixels", ...)
    }, error = function(e) {
      # Fallback: if auto-bandwidth fails, try a simple rule or fixed sigma
      # Adjust or expose this fallback based on your data scale
      density.ppp(pp, at="pixels", sigma = bw.ppl(pp))
    })
    
    xy <- coords(pp)
    pint <- lookup.im(kern, xy$x, xy$y)
    
    good <- which(is.finite(pint) & !is.na(pint) & pint > 0)
    if (length(good) == 0) return(pp[integer(0)])
    if (length(good) <= n_target) return(pp[good])
    
    w <- 1 / pint[good]
    w <- w / sum(w)
    
    # Defensive programming: use a robust sampling wrapper to avoid the sample(x) gotcha
    sampled_idx <- if (length(good) == 1) good else sample(good, n_target, replace = FALSE, prob = w)
    
    return(pp[sampled_idx])
    
  }, mc.cores = min(length(pp_split), detectCores()))
  
  # 4. Recombine using the explicit geometry package namespace
  pp_final <- do.call(superimpose, sampled_list)
  
  return(pp_final)
}
```

Creation of the subsetted pp object

```{r}
#| label: pp-subsetting

# Define the distance vector
r_values <- seq(0, 150, length.out = 70)

# Reproducibility for downsampling
set.seed(10)
ppp_sub <- downsample_KDE(pp_object, c("neoplastic", "stromal"), 40000)
```

Estimating the Lcross function between the two cell types with the subsetted pp object

```{r}
# Compute Lcross function
Lcross_neoStroma <- Lcross(
  ppp_sub, 
  i = "neoplastic", 
  j = "stromal", 
  r = r_values,
  correction = "border"
)
```

## Saving Results

Export the resulting analysis object into a serialized `.rds` format.

```{r}
#| label: save-output
#| eval: FALSE

saveRDS(Lcross_neoStroma, file = save_path)
```

